---
title: "Alpha-beta pruning"
author: "Alex Zobolev"
date: today
subtitle: "Preliminary results"
abstract: ""
execute:
  echo: false
  warning: false
  message: false
  eval: true
params:
    i_norm_counts_tab: "results/ASVs_counts_normalized.tsv"
    i_distances_matrix: "results/ASVs_euclidean_distance.tsv"
    i_taxa_table: "results/ASVs_taxa.tsv"
    all_meta: "resources/metadata/all_meta.csv"
---

```{r}
#| output: false
#| label: prepare data

library(phyloseq)
library(ggplot2)
library(vegan)
library(gt)
library(dplyr)
library(tibble)
library(dendextend)

source("workflow/scripts/plots/pcoa.R")
source("workflow/scripts/plots/dendrogram.R")


norm_counts_tab <- read.csv(params$i_norm_counts_tab,sep = "\t", row.names='X')
samples_info <- read.csv(params$all_meta, row.names='Run')[colnames(norm_counts_tab), ]
euc_dist_all <- as.dist(as.matrix(read.csv(
    params$i_distances_matrix, 
    sep='\t', 
    header=T,
    row.names=1
)))
taxa_table <- as.matrix(read.csv(params$i_taxa_table, sep = "\t", row.names=1))


counts_phy <- otu_table(norm_counts_tab, taxa_are_rows = T)
samples_info_phy <- sample_data(samples_info)
taxa_table_phy <- tax_table(taxa_table)

phy <- phyloseq(counts_phy, samples_info_phy, taxa_table_phy)

meta <- data.frame(sample_data(phy))
predictors <- c('depr_or_control', 'batch', 'd_sex', 'd_age_hc')
predictors <- predictors[meta[predictors] |> sapply(function(x) length(unique(x))) > 1]
rhs <- paste(predictors, collapse = " + ")
```

# Tree of samples

```{r}
#| label: tree-samples
#| fig-width: 10
#| fig-height: 10

plot_dendrogram(euc_dist_all, phy, depr_or_control)
```

As we can see, there is no clear separation between groups of samples. There are some clusters of samples, but they are not related to the groups of interest (depression vs control).

# Beta dispersion

## Plot ordination using PCoA

Plotting PCoA with Euclidean distance, we can see that there is no clear separation between groups.

```{r}
#| label: ord-plot
pcoa <- plot_pcoa(phy, euc_dist_all, depr_or_control)$pcoa
plot_pcoa(phy, euc_dist_all, paste(depr_or_control, season, sep="_"))
``` 

```{r}
#| label: find-pcoa-outliers
scores <- as.data.frame(pcoa$points)
d2 <- mahalanobis(
    scores,
    colMeans(scores),
    cov(scores)
)
outliers <- which(
    d2 > qchisq(0.975, df = 2)
)
```

There some extremes samples (`{r} names(outliers)`) that are far from the others. So i have to check what is biological or technical reason for it.

Then, I'll check where on tree are these samples

```{r}
#| label: tree-extremes
#| fig-width: 10
#| fig-height: 10

plot_dendrogram(euc_dist_all, phy, rownames(sample_data(phy)) %in% names(outliers))
```

These samples located on different leafs of dendrogram

## PERMANOVA

I use PERMANOVA (Permutational Multivariate Analysis of Variance) to test Beta diversity.

```{r}
pmnv_test <- function (frm) {
    result <- adonis2(frm, data = meta, by='term', parallel = 30)
    formatted_result <- result |>
        as.data.frame() |>
        tibble::rownames_to_column("term") |>
        gt(rowname_col = "term") |>
        gt::fmt_number(columns = c("R2", "F", "Pr(>F)"), decimals = 4)
    return(list(
        result=result,
        formatted_result=formatted_result
    ))
}
form_string <- paste("euc_dist_all", "~", rhs)
f <- as.formula(form_string)
```



Checking Beta dispersion between Microbiota of pations with depression and without it using PERMANOVA test with formula `{r} form_string`.

```{r}
#| label: permanova-depression
#| tbl-cap: "Beta dispersion between Microbiota of pations with depression and without it using PERMANOVA test"
depr_vs_control_pmnv <- pmnv_test(f)
depr_vs_control_pmnv$formatted_result
```

So there is significant (p-value=`{r} depr_vs_control_pmnv$result[1, 'Pr(>F)']`) beta dispersion between Microbiota of pations with depression. Microbiota features explains `{r} depr_vs_control_pmnv$result[1, 'R2']` of differences between patients depression status.

