---
title: "Alpha/Beta"
author: "Alex Zobolev"
date: today
subtitle: "Preliminary results"
abstract: ""
execute:
  echo: false
  warning: false
  message: false
  eval: true
params:
    i_counts_tab: "results/ASVs_counts.tsv"
    i_norm_counts_tab: "results/ASVs_counts_normalized.tsv"
    i_distances_matrix: "results/ASVs_euclidean_distance.tsv"
    i_taxa_table: "results/ASVs_taxa.tsv"
    i_reads_loss: "results/reads_loss_summary.tsv"
    all_meta: "resources/metadata/all_meta.csv"
---

```{r}
#| output: false
#| label: prepare data

library(phyloseq)
library(ggplot2)
library(vegan)
library(gt)
library(dplyr)
library(tibble)
library(dendextend)
library(dendextend)
library(ggpubr)
library(patchwork)

source("workflow/scripts/plots/pcoa.R")
source("workflow/scripts/plots/dendrogram.R")
source("workflow/scripts/plots/rarecurve.R")
source('workflow/scripts/plots/alpha_violins.R')


norm_counts_tab <- read.csv(params$i_norm_counts_tab,sep = "\t", row.names='X')
counts_tab <- read.csv(params$i_counts_tab,sep = "\t", row.names='X')

samples_info <- read.csv(params$all_meta, row.names='Run')[colnames(norm_counts_tab), ]
euc_dist_all <- as.dist(as.matrix(read.csv(
    params$i_distances_matrix, 
    sep='\t', 
    header=T,
    row.names=1
)))
taxa_table <- as.matrix(read.csv(params$i_taxa_table, sep = "\t", row.names=1))


samples_info_phy <- sample_data(samples_info)
taxa_table_phy <- tax_table(taxa_table)

phy_norm <- phyloseq(otu_table(norm_counts_tab, taxa_are_rows = T), samples_info_phy, taxa_table_phy)
phy_raw <- phyloseq(otu_table(counts_tab, taxa_are_rows = T), samples_info_phy, taxa_table_phy)

meta <- data.frame(sample_data(phy_norm))
predictors <- c('depr_or_control', 'batch', 'd_sex', 'd_age_hc')
predictors <- predictors[meta[predictors] |> sapply(function(x) length(unique(x))) > 1]
rhs <- paste(predictors, collapse = " + ")
```

# Tree of samples

```{r}
#| label: tree-samples
#| fig-width: 10
#| fig-height: 10

plot_dendrogram(euc_dist_all, phy_norm, depr_or_control)
```

As we can see, there is no clear separation between groups of samples. There are some clusters of samples, but they are not related to the groups of interest (depression vs control).

# Beta dispersion

## Depression vs. control

Plotting PCoA with DESeq2 VST normalization and Euclidean distance, we can see that there is no clear separation between groups.

```{r}
#| label: ord-plot
#| fig-cap: "Plot ordination plot by ASV for Depression vs. Control"
ord_res <- plot_pcoa(phy_norm, euc_dist_all, depr_or_control, legend_title="Depression vs. Control")
pcoa <- ord_res$pcoa
``` 

Firstly we tested dispersions heterogeneous with `betadisper` form `vegan` package

```{r}
#| label: dc-betadisper
anova(betadisper(euc_dist_all, sample_data(phy_norm)$depr_or_control), by='margin')

betadisper(euc_dist_all, sample_data(phy_norm)$depr_or_control)$group.distances
```

Depression and control samples have heterogeneous dispersions (spread of control samples twice bigger then depression set), which means I cannot truly trust permanova test on whole set. But amount of samples are the same for depression and control sets. So premanova still valid choice. 

Associations between microbial taxa and depression of host were identified using permutational multivariate analysis of variance (PERMANOVA). The adonis2 function from the vegan package was employed, with permutation tests performed with 1 000 permutations.

```{r}
#| label: dc-premanova
#| tbl-cap: "Results of permutational multivariate analysis of variance. There is significant difference between beta dispersion of depression and control sets"

form_string <- paste("euc_dist_all", "~", rhs)
f <- as.formula(form_string)
adonis2(f, meta, by = 'margin', permutations = 999)
```

Differences of dispersion are illustrated by different size of groups ellipses on PCoA plot. And differences of centroids - by gap between ellipses. 

### Interim conclusions: Beta

__Significant differences in gut microbiome composition were detected between patients with depression and the control group. These differences are associated with variations in the mean community structure and decreased microbial community dispersion in depressed patients.__

## Seasonal depression

```{r}
#| label: sd-pcoa
#| fig-cap: "Plot ordination plot by ASV for seasonal depression"
plot_pcoa(phy_norm, euc_dist_all, paste(depr_or_control, season, sep="_"), legend_title="Depression vs. Season")
```

```{r}
#| label: sd-betadisper
anova(betadisper(euc_dist_all, paste(sample_data(phy_norm)$depr_or_control, sample_data(phy_norm)$season, sep="_")), by='margin')
```

```{r}
#| label: sd-premanova
#| tbl-cap: "Results of permutational multivariate analysis of variance. There is significant connection between microbial taxa and seasonal depression"
form_string <- paste("euc_dist_all", "~", "season * ", rhs)
f <- as.formula(form_string)
adonis2(f, meta, permutations = 999)
```


# Alpha

```{r}
#| label: rarecurve
#| fig-cap: "Rarefaction curves for individual samples"
plot_rarecurve(phy_raw, depr_or_control)
```

The diversity composition of the bacterial microbiota was assessed using alpha diversity (α-diversity), specifically the Shannon and Simpson indexes at ASV levels. Function `ggviolin` form package `ggpubr` was used to draw Shannon and Simpson indexes distributions across samples. Function `stat_compare_means` with method `t.test` form the same package was used to compare these distributions.

```{r}
#| label: alpha-metrics
#| fig-cap: "Alpha-Diversity Analysis. Shannon, Simpson, and Chao's indices were used"
phy_raw_no_zero_taxa <- prune_taxa(taxa_sums(phy_raw) > 0, phy_raw)
alpha <- estimate_richness(phy_raw_no_zero_taxa, measures = c("Shannon", "Simpson", "Chao1"))
plot_alpha_violins(cbind(sample_data(phy_raw), alpha), "depr_or_control",  c("Shannon", "Simpson"), 'results/figures/alpha.png')
```

As we can see, there are no significant difference of alpha indexes distributions (p-value > 0.5 for both indexes) 